\section{Algorithms}
In this section several algorithms are described, which all try to learn the parameters of a \gls{hmm} given a set of training sequences $D$ and a set of validation sequences $V$, which contain a number of sequences over $m$ distinct symbols.
In other words, each algorithm will try to find a model that makes the generation of the particular validation sequences most likely.
Since a \gls{hmm} can be represented by the use of either matrices or a graph-like structure, any of the algorithms may output either a matrix or a graph representation of a \gls{hmm}, denoted by $M$ or $G$ respectively.
By $LL(M)$ or $LL(G)$ we denote the logarithmic likelihood of the validation sequences given the model.ss

The algorithms proposed in this section have been divided into two types.
The static algorithms start with a \gls{hmm} with a particular number of states which will never change.
The dynamic algorithms start with a small \gls{hmm} in regards to the number of states, and will dynamically extend the number of states through a number of iterations. 
For the static algorithms, $n$ denotes the number of states to be used.
For the dynamic algorithms, $n$ denotes a number of iterations $1, ..., n$, where each iteration extends the model by a single node.
In \ref{fig:alg-hierarchy} the names and classes of our proposed algorithms can be seen. Note that the \gls{baum-welch} has also been included, simply to emphasize which class of algorithms it belongs to. 

\begin{figure}[!h]
\Tree[.Algorithms
		[.{Static size} 
			{Baum-Welch}
            {Sparse Baum-Welch}
        ]
       	[.{Dynamic size} 
       		{Greedy Extend}
       		{State Splitting}
      	]
     ]
\caption{The algorithms used in our experiments}
\label{fig:alg-hierarchy}
\end{figure}

Since the \gls{baum-welch} is guaranteed to never worsen $LL(G)$ when run on $G$ (or $LL(M)$ on $M$ respectively), it is used internally by most of the algorithms.
By $BW_t(M, D, V)$ we denote the \gls{hmm} obtained after running Baum Welch on the \gls{hmm} $M$ using the training sequences $D$ and validation sequences $V$, and iterating as long as each iteration increases the logarithmic likelihood of the validation sequences by at least $t$.
By $BW^i(M, D)$ we denote the \gls{hmm} obtained after running $i$ iterations of Baum Welch on the \gls{hmm} $M$ using the training sequences $D$.
If a \gls{hmm} is represented by a graph $G$, respective notations $BW_t(G, D, V)$ and $BW^i(G, D)$ are used.

\input{./content/Models/Algorithms/sparsebaumwelch}
\input{./content/Models/Algorithms/greedyextend}
\input{./content/Models/Algorithms/statesplitting}
%\input{./content/Models/Algorithms/gammastatesplitter}
