\subsection{Algorithms}
In this section a number of algorithms are described, which all try to learn the parameters of a \gls{hmm} given a set of training sequences $D$, which contains a number of sequences over $s$ distinct symbols.
In other words, each algorithm will try to find a model that makes the generation of the particular training sequences most likely.
Since a \gls{hmm} can be represented by the use of either matrices or a graph like structure, any of the algorithms may output either a matrix or graph representation of a \gls{hmm}, denoted by $M$ or $G$ respectively.
By $LL(M)$ or $LL(G)$ we denote the likelihood of the training sequences given the model.

The algorithms proposed in this section have been divided into two types.
The static algorithms start with a \gls{hmm} with a particular number of states which will never change.
The dynamic algorithms start with a \gls{hmm} of just a single state, and will dynamically extend the number of states through a number of iterations. 
For the static algorithms, $n$ denotes the number of states to be used.
The dynamical algorithms will by $n$ denote a number of iterations $1, ..., n$, where each iteration extends the model by a single node.

Since the Baum Welch algorithm is guaranteed to never worsen $LL(G)$ of any $G$, it is natural that the Baum Welch is used internally by many of the proposed algorithms.
By $BW_t(M, D)$ we denote the \gls{hmm} obtained after running Baum Welch on the \gls{hmm} $M$ using the training sequences $D$, and iterating as long as each iteration increases the likelihood by at least $t$.
By $BW^i(M, D)$ we denote the \gls{hmm} obtained after running $i$ iterations of Baum Welch on the \gls{hmm} $M$ using the training sequences $D$.
If a \gls{hmm} is represented by a graph, we denote it $G$, and the similar notations $BW_t(G, D)$ and $BW^i(G, D)$ are used.
\Tree[.Algorithms
		[.{Static size} 
			{Baum Welch}
            {Sparse Baum Welch}
        ]
       	[.{Dynamic size} 
       		{Strict George}
       		{George}
       		{Greedy Extend}
      	]
     ]

\input{./content/Experiments/Algorithms/greedyextend}
\input{./content/Experiments/Algorithms/sparsebaumwelch}
\input{./content/Experiments/Algorithms/statesplitting}