\chapter{Experiments}
\label{chap:experiments}
The goal for the experiments is to discover strength and weaknesses for our algorithms and models, given data created by different machines. In section \ref{sec:testenv} the architecture of the used experiment environment is described. In section \ref{sec:pautomac} it is described how Pautomac used 4 parameters for their machines: the state space, the transition sparsity, the emission symbol sparsity and the amount of symbols used in the model. We expect these parameters to produce learning complexity in different ways, where we wish to illustrate how 3 of these 4 parameters effects each algorithm and model, as well as observe an overall logarithmic likelihood score. In an attempt to save time and still produce concise results, we have chosen to use 11 data sets from Pautomac, how these datasets were chosen is described in section \ref{sec:datasets}. In section \ref{sec:parameters} we have studied what parameters the experiments should by run with, including the amount of training data used, \gls{bw}'s threshold and number of \gls{bw} iterations for the \gls{ge} and \gls{gs} algorithm. In section \ref{sec:greedy} we define how many iterations of \gls{bw} that is needed to support strong learning without spending unnecessary amounts of time.
Finally in section \ref{sec:results} the experiments for each parameter experiment can be found. A couple concrete Pautomac competition "submission attempt" has been added as well, to show how our work relates to some of the best results from the competition. In the last section a benchmark of the running time of some algorithms is presented.

\input{./content/Experiments/choice-of-programming-language.tex}
\input{./content/Experiments/test-environment.tex}
\input{./content/Experiments/datasets.tex}
\input{./content/Experiments/parameters.tex}
\input{./content/Experiments/experiment-description.tex}
\input{./content/Experiments/speedrun.tex}

\FloatBarrier

\paragraph{State Space Experiments}

The purpose of the state experiments are to show how \gls{bw} and our own algorithms and models perform in relation to different number of states.\\

	\input{./content/Experiments/graphdata/state-result.tex}

Initial observations
\begin{itemize}
\item GE performs the best, except dataset 1. It seems like GE would surpass BW with a larger state space.
\item SBW performs worse than BW, note that dataset 41 has a fairly small y-axis, where the best GE result is about 800 points. We estimate an error margin of 100-200 points. Without more runs for each experiment, it is impossible to determine an exact error margin.
\item In general we seem to 
\end{itemize}


\paragraph{Transition Density Experiments}

The purpose of the transition density experiments are to show how \gls{bw} and our own algorithms and models perform in relation to the transition density.\\

	\input{./content/Experiments/graphdata/density-result.tex}

Initial observations
\begin{itemize}
\item GE performs the best
\item 
\item 
\end{itemize}

\paragraph{Symbol Alphabet Size Experiments}
The purpose of the s are to show how \gls{bw} and our own algorithms and models perform in relation to the size of the emission symbol alphabet size.\\

	\input{./content/Experiments/graphdata/symbol-result.tex}
	
Initial observations
\begin{itemize}
\item BW performs the best on 3 out of 4 sets
\item 
\item 
\end{itemize}	
	
%\subsection{Greedy Extend Experiments}
	%\input{./content/Experiments/results/greedyextend-results.tex}
