\FloatBarrier

\section{Experiments}\label{sec:results}

\subsection{Dataset Parameter Experiment}
The approach used for each dataset parameter experiment have been to cover a range of state amounts, ranging from 10 to 100 states, together with the previously define parameters. \gls{bw} was probed on specific state amounts with step sizes of 10, where the dynamic algorithms have been defined to reach either some local maxima or stop at 100 states.

\paragraph{State Space Experiments}
The purpose of the state experiments are to show how \gls{bw} and our own algorithms and models perform in relation to the state space.

	\input{./content/Experiments/graphdata/state-result.tex}

Initial observations
\begin{itemize}
\item GE performs the best, except dataset 1. It seems like GE would surpass BW with a larger state space.
\item SBW performs worse than BW, note that dataset 41 has a fairly small y-axis, where the best GE result is about 800 points. We estimate an error margin of 100-200 points. Without more runs for each experiment, it is impossible to determine an exact error margin.
\item In general we seem to 
\end{itemize}


\paragraph{Transition Density Experiments}
The purpose of the transition density experiments are to show how \gls{bw} and our own algorithms and models perform in relation to the transition density.

	\input{./content/Experiments/graphdata/density-result.tex}

Initial observations
\begin{itemize}
\item GE performs the best
\item 
\item 
\end{itemize}

\paragraph{Symbol Alphabet Size Experiments}
The purpose of the s are to show how \gls{bw} and our own algorithms and models perform in relation to the size of the emission symbol alphabet size.

	\input{./content/Experiments/graphdata/symbol-result.tex}
	
Initial observations
\begin{itemize}
\item BW performs the best on 3 out of 4 sets
\item 
\item 
\end{itemize}	
	
%\subsection{Greedy Extend Experiments}
	%\input{./content/Experiments/results/greedyextend-results.tex}