The first big question about the Greedy Extend algorithm, is how the choice of $\beta$ affects the performance of the algorithm.
As $\beta$ denotes the number of Baum Welch iterations to run each time the algorithm attempts to extend the graph, increasing $\beta$ will also increase the run time of the algorithm. It may be the case that better results are achieved when $\beta$ is increased, since more iterations of Baum Welch also means a greater increase in likelihood. However, it could be the case that using many iterations early increases the chance of getting trapped in a local optimum.
An experiment has been conducted of using different values for $\beta$ on data set $1$ from the Pautomac competition. The results can be seen in figure \ref{fig:ge-different-thresholds-tested}. Each line represents the mean value of 5 runs of the Greedy Extend algorithm with the specified number of iterations. Some of the plots have been cut off at a point where one of the runs did not manage to extend beyond a certain number of states.

\begin{figure}
\begin{centering}
\begin{tikzpicture}
	\pgfplotsset{every axis legend/.append style={ 
		at={(0.5,1.03)},
		anchor=south}}
	\begin{axis}[
			xlabel = Number of states,
            	ylabel = Score (lower is better),
            	legend columns=-1,
            	legend entries={IT-0, IT-1, IT-2, IT-3, IT-5, IT-10, IT-50},
			legend style={/tikz/every even column/.append style={column sep=0.3cm}}]
		
		\addplot+table[x=States, y=IT-0, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
		
		\addplot+table[x=States, y=IT-1, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
		
		\addplot+table[x=States, y=IT-2, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
		
		\addplot+table[x=States, y=IT-3, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};

		\addplot+table[x=States, y=IT-5, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
		
		\addplot+table[x=States, y=IT-10, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
		
		\addplot+table[x=States, y=IT-50, col sep=tab]
		{content/Experiments/graphdata/ge-intermediate-iterations-test.csv};
	\end{axis}
\end{tikzpicture} 
\caption{Test of different values for $\beta$ while running the Greedy Extend algorithm.}
\label{fig:ge-different-thresholds-tested} 
\end{centering}
\end{figure}

The figure shows surprisingly that using no iterations is somehow better than using just a single iteration. However, using 5 or more iterations is better than not using any iterations at all.
When using 5 iterations, only a single run did not reach 50 states (it stopped at 49).
We choose to conduct further experiments with $\beta = 10$, since all runs with 10 iterations reached 50 states, and it looks like the performance does not change significantly when increasing the number of iterations beyond 10.